Hydraulically-vulnerable trees are buffered by deep-water access in a tropical forest community

### 2020-10-24 manuscript.rmd
1. added a comment on impact of stem water storage to ERD in M&M

To estimate effective rooting depth, we tested alternate models for growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many intrinsic and extrinsic factors that limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Rhizosphere soil water potential ($\Psi$~soil~) down-regulates plant tissue water potentials [@Christoffersen.2018]. In moist forests $\Psi$~leaf~ equilibriates with $\Psi$~soil~ at night and declines through the day as plants transpire. $\Psi$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which in turn controls stomatal conductance and photosynthetic rates [@Sack.2012]. We therefore used maximum diurnal leaf conductance ($K_\textrm{leaf}$) as determined by $\Psi$~soil~ as a proxy for the soil water limitation on growth. In our ERD models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential. This approach ignores the buffering effect of stem water storage on $K_\textrm{leaf}$, which may be a subject of future work (see Discussion). VPD has a positive impact on transpiration and thus photosynthesis and growth upto a VPD threshold, after which VPD has a negative impact [@Grossiord.2020]. The fraction of a tree's leaf cover present at a given time further limits it's growth potential. We therefore regressed species-specific growth in a census interval against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

>>>>

To estimate effective rooting depth, we tested alternate models for growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many intrinsic and extrinsic factors that limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Rhizosphere soil water potential ($\Psi$~soil~) down-regulates plant tissue water potentials [@Christoffersen.2018]. In moist forests $\Psi$~leaf~ equilibriates with $\Psi$~soil~ at night and declines through the day as plants transpire. $\Psi$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which in turn controls stomatal conductance and photosynthetic rates [@Sack.2012]. We therefore used maximum diurnal leaf conductance ($K_\textrm{leaf}$) as determined by $\Psi$~soil~ as a proxy for the soil water limitation on growth. In our ERD models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential. This approach ignores the buffering effect of stem water storage on $K_\textrm{leaf}$, which may cause an assignment of a slightly deeper ERD to a species. Future work should therefore explore the influence of stem water storage on estimation of ERD (see Discussion). VPD has a positive impact on transpiration and thus photosynthesis and growth upto a VPD threshold, after which VPD has a negative impact [@Grossiord.2020]. The fraction of a tree's leaf cover present at a given time further limits it's growth potential. We therefore regressed species-specific growth in a census interval against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

### 2020-10-24 old_ms_sections.Rmd
1. added a comment on impact of stem water storage to ERD in M&M

@@

@@

Deeper ERD species had on average more hydraulically efficient, but risky, suite of traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk. By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\Psi_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$. Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period. @Condit.2017 showed that half of the extreme dry seasons (< 1 SD annual water deficit) during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015).

>>>>

## Intro

Historic climatic variation is relatively small (Mora et al., 2013), and, though data have been scant, trees' hydraulic safety margins are reportedly relatively narrow (Choat et al., 2012; but see Oliveira et al., 2019).

## M&M

@@

## Results

The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_\textrm{max,stem}$) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016]. We report such a trade-off ($n$ = 26 species) at our study site stem level.

@@

Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period. @Condit.2017 showed that half of the extreme dry seasons (< 1 SD annual water deficit) during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015).
